---
title: "Data Analysis Project"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "")
```



# Student Information

**Name**: Noel Joseph Philip
**G Number**: G01323826




# Introduction

This semester we will be working with a dataset of all domestic outbound flights from Dulles International Airport in 2016.

Airports depend on accurate flight departure and arrival estimates to maintain operations, profitability, customer satisfaction, and compliance with state and federal laws. Flight performance, including departure and arrival delays must be monitored, submitted to the Federal Aviation Agency (FAA) on a regular basis, and minimized to maintain airport operations. **The FAA considered a flight to be delayed if it has an arrival delay of at least 15 minutes**.

The executives at Dulles International Airport have hired you as a Data Science consultant to perform an exploratory data analysis on all domestic flights from 2016 and produce an executive summary of your key insights and recommendations to the executive team.

Before you begin, take a moment to read through the following airline flight terminology to familiarize yourself with the industry: [Airline Flight Terms](https://www.myairlineflight.com)



## Dulles Flights Data

The `flights_df` data frame is loaded below and consists of 33,433 flights from IAD (Dulles International) in 2016. The rows in this data frame represent a single flight with all of the associated features that are displayed in the table below.

**Note**: If you have not installed the `tidyverse` package, please do so by going to the `Packages` tab in the lower right section of RStudio, select the `Install` button and type `tidyverse` into the prompt. If you cannot load the data, then try downloading the latest version of R (at least 4.0). The `readRDS()` function has different behavior in older versions of `R` and may cause loading issues.



```{r}

library(tidyverse)

flights_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/dulles_flights.rds'))

```


### Raw Data

```{r}

flights_df

```


# Exploratory Data Analysis


Executives at this company have hired you as a data science consultant to evaluate their flight data and make recommendations on flight operations and strategies for minimizing flight delays.

You must think of **at least 8 relevant questions** that will provide evidence for your recommendations.

The goal of your analysis should be discovering which variables drive the differences between flights that are early/on-time vs. flights that are delayed.

Some of the many questions you can explore include:

-   Are flight delays affected by taxi-out time? Do certain airlines or time of 
    year lead to greater taxi out times (i.e. traffic jams on the runways)?

-   Are certain times of the day or year problematic?

-   Are certain destination or airlines prone to delays?


You must answer each question and provide supporting data summaries with either a summary data frame (using `dplyr`/`tidyr`) or a plot (using `ggplot`) or both.

In total, you must have a minimum of 5 plots and 4 summary data frames for the exploratory data analysis section. Among the plots you produce, you must have at least 4 different types (ex. box plot, bar chart,
histogram, heat map, etc...)

Each question must be answered with supporting evidence from your tables and plots.


## Question 1



**Question**: Does the destination airport state of the flight trip impact the arrival delay time?



**Answer**:



To add additional R code chunks for your work, select `Insert` then `R` from the top of this notebook file.

```{r}
flights_df %>%
  filter(arrival_delay > 15) %>%
  ggplot(aes(x = fct_infreq(dest_airport_state)))+
    geom_bar()+theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



```




## Question 2



**Question**: Do destination region (South, West, Midwest, Northeast etc.) of the flight trip affects the arrival delay time?



**Answer**:




```{r}

flights_df %>% filter(
  arrival_delay > 15
) %>%
  group_by(dest_airport_region) %>% 
                  summarise(n_flights = n(),
                            min_delay = min(arrival_delay),
                            avg_delay = mean(arrival_delay),
                            max_delay = max(arrival_delay),
                            sd_delay = sd(arrival_delay))


```



## Question 3



**Question**: Is the flight distance and arrival delay at destination report related?



**Answer**:




```{r}
flights_df %>% 
  mutate(distance = cut(flights_df$distance,
                        breaks=c(0, 700, 3000, 5000),
                        labels=c('Short-haul', 'Medium-haul', 'Long-haul'))) %>% 
  group_by(distance) %>% 
  summarise(no_of_flights = n())

```

```{r}
flights_df %>% filter(
  arrival_delay > 15
) %>% 
  mutate(distance = cut(distance,
                        breaks=c(0, 700, 3000, 5000),
                        labels=c('Short-haul', 'Medium-haul', 'Long-haul'))) %>%
  ggplot(aes( x = distance, y = arrival_delay, fill = distance))+
  geom_boxplot()
  
```



## Question 4



**Question**: Do some airlines tend to cause delay at the destination airport?



**Answer**:




```{r}
flights_df %>%
  filter (
    arrival_delay > 15
  ) %>%
  ggplot(aes(x = arrival_delay, fill = airline))+
  geom_histogram()+
  facet_wrap(~ airline, nrow = 5)



```
```{r}
flights_df %>%
  filter (
    arrival_delay > 15
  ) %>%
  group_by(airline) %>%
  summarise(n_flights = n(),
                            min_delay = min(arrival_delay),
                            avg_delay = mean(arrival_delay),
                            max_delay = max(arrival_delay),
                            sd_delay = sd(arrival_delay))
```




## Question 5



**Question**: Does Taxi In time cause flight to delay after reaching the airport.



**Answer**:




```{r}
flights_df %>%
  filter(
    arrival_delay > 15
  ) %>%
  ggplot(aes(x = taxi_out, y = dep_delay))+
  geom_point()



```


## Question 6



**Question**: 



**Answer**:




```{r}

flights_df %>% 
  filter(arrival_delay > 15) %>%
  mutate(sch_arrival_timeslot = cut(sch_arrival_time,
                            breaks = c(0.00,7.00,12.00,16.00,20.00,24.00),
                            labels = c('Early Morning', 'Morning', 'Mid-day', 'Evening', 'Night'))) %>%
  group_by(sch_arrival_timeslot) %>%
  summarise( no_of_delayed_flights_arrived = n())
  ggplot(aes())


```
```{r}
flights_df %>%
  mutate(sch_arrival_timeslot = cut(sch_arrival_time,
                            breaks = c(0.00,7.00,12.00,16.00,20.00,24.00),
                            labels = c('Early Morning', 'Morning', 'Mid-day', 'Evening', 'Night'))) %>%
  mutate( delayed = cut(arrival_delay,
                        breaks = c(0,15,max(arrival_delay)),
                        labels = c('Not Delayed', 'Delayed'))) %>%
  ggplot(aes(x = sch_arrival_timeslot, y = arrival_delay))+
  geom_boxplot()
```


## Question 7



**Question**: Do weekends causes more delays at destination & arrival airports compared to week days.



**Answer**:




```{r}
flights_df %>% 
  filter( arrival_delay > 15) %>%
  mutate( weekend = case_when(weekday == 'Saturday' | weekday == 'Sunday' ~ 'Yes',
                              TRUE ~ 'No')) %>%
  group_by(weekend) %>%
  summarise(delayed_flights = n(),
            average_delay = mean(arrival_delay))



```

```{r}
flights_df %>% 
  mutate( weekend = case_when(weekday == 'Saturday' | weekday == 'Sunday' ~ 'Yes',
                              TRUE ~ 'No')) %>%
  group_by(weekend) %>%
  summarise(arrived_flights = n(),
            average_delay = mean(arrival_delay))
```
```{r}
flights_df %>% 
  filter( arrival_delay > 15) %>%
  mutate( weekend = case_when(weekday == 'Saturday' | weekday == 'Sunday' ~ 'Yes',
                              TRUE ~ 'No')) %>%
  ggplot(aes(x = dep_delay, y = arrival_delay, color = weekend))+
  geom_point()+
  facet_wrap(~ weekend, ncol = 1)
```


## Question 8



**Question**:



**Answer**:




```{r}

flights_df %>% 
  filter( arrival_delay > 15) %>%
  group_by(scheduled_flight_date) %>%
  summarise(total = n()) %>%
  ggplot(aes(x = scheduled_flight_date, y = total)) +
  geom_line()


```


# Summary of Results

Write an executive summary of your overall findings and recommendations to the executives at Dulles Airport. Think of this section as your closing remarks of a presentation, where you summarize your key findings and make recommendations on flight operations and strategies for minimizing flight delays.

Your executive summary must be written in a [professional tone](https://www.universalclass.com/articles/writing/business-writing/appropriate-tone-in-business-communications.htm), with minimal grammatical errors, and should include the following
sections:

1.  An introduction where you explain the business problem and goals of your data analysis

    -   What problem(s) is this company trying to solve? Why are they important
        to their future success?

    -   What was the goal of your analysis? What questions were you 
        trying to answer and why do they matter?


2.  Highlights and key findings from your Exploratory Data Analysis section

    -   What were the interesting findings from your analysis and
        **why are they important for the business**?

    -   This section is meant to **establish the need for your recommendations** 
        in the following section
        

3.  Your recommendations to the company

    -   Each recommendation must be supported by your data analysis results

    -   You must clearly explain **why** you are making each recommendation and which results 
        from your data analysis support this recommendation

    -   You must also describe the potential business impact of your recommendation:

        -   Why is this a good recommendation?

        -   What benefits will the business achieve?
      


## Executive Summary

Introduction:
The Dulles International Airport is one of the busiest airports in the country. It services both international and domestic passengers who travels within and outside to the country. As a result of its proximity to the DC metropolitan area it usually has great numbers of people, workers and businessmen who commute via the airport to other major cities and states. This causes the airport to service a lot of people at the same time which leads to decrease in the quality of services and other issues. 
One of the biggest challenges that had risen in the past couple of years is the problem of delayed arrival of flights at the destination airport frequently. The goal of this analysis is to find out the factors that causes this delay of flights at the destination airport.  Using the analysis, some recommendations can be formulated that can help resolve this problem.
Key Analysis and Interpretations:
The Dulles International airport is servicing flights to all the important cities across the country which points to our first key finding, i.e., airport that reports the most numbers of delayed flights from Dulles airport. Form the analysis, we can see that the California airports were the airports with highest numbers of reported delayed flights. This is followed by airports in Florida, Colorado, Texas, Georgia and so on. Further, an analysis of the major geographic regions where each of these flights arrive shows that flights headed towards the west suffer more delays compared to the other regions of the US. However, it is to be noted that the average delay in minutes is highest among the flights headed to the Midwest as compared to that to the west.
The above-mentioned analysis evidently shows that flights to the west especially those headed to airports in the state of California suffers delayed arrival most frequently. Since flights to the west usually takes more duration than flights to anywhere else, there is a need to investigate the impact of distance in delaying the flights’ arrival. The analysis was conducted by splitting the flight trips into three groups based on the distance of the trip; they are short-haul, medium-haul and long-haul. The analysis shows that long-haul flights are bound to delay more frequently than short-haul and medium haul flights.
The next analysis focuses on the various airline services that operates flights from Dulles to other airports. Currently, there are eight different airline services that are operating in Dulles airport. The analysis reveled that almost all airlines suffer from arrival delay. United Airlines operates the greatest number of domestic flights from Dulles; hence their airlines have the greatest number of delayed flights. Among the different flight operators, the SkyWest airline is the airline with the highest average delay followed by ExpressJet.
The next analysis considers the delayed flights’ taxi out time and departure delays. From the analysis we can confirm that departure delays significantly impact the arrival delays. On the contrary, flight taxi out time do not impact either the departure or arrival delays. This points to the fact that most of the delays occurs due to the passenger onboarding procedures at the airport.
Another important factor is the time of arrival of flights, which can be categorized as early morning flights (12 am to 7 am), morning flights (8 am to 12 pm), mid-day flights (1pm to 4 pm), evening flights (5pm to 8pm) and night flights (8pm to 11.59pm). However, after excluding the outliers the flight delays are similar in each time categories. 
From the summary of the next analysis, it is evident that there is relatively a smaller number of flights being operated from the airport during the weekends as compared to weekdays. However, the percentage of flights delayed during weekends is comparatively higher than that during the weekdays. This may be due to a smaller number of staff members or employees at the airport to deal with the sudden rush of passengers on weekends.
Finally, the analysis concludes by looking at the overall number of delayed flights reported between January 2016 and January 2017. The line graph shows the increasing trend of the delayed arrival over the year. A seasonal hike can be witnessed during the months of June, July and August which are the summer vacation months.
Conclusion:
From all the analysis that were conducted, it seems that certain ground level changes are required at the Dulles International Airport to better service its passengers. Firstly, the airport needs to have a separate crew for conducting passenger onboarding for travelers heading to states and cities in the west. This will not only reduce the chance of arrival delay of long-haul trips, but also reduces the rush at the airport which makes it easier for conducting on-boarding for travelers heading to other directions as well.
Also, there needs to be an effective way to handle the rush that may happen during the weekends. This can be done either by employing more crew members for weekend shifts or by implementing automated passenger onboarding kiosks that streamlines passenger requests.
















